%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}
	
I would like to express my sincere gratitude to everyone that supported me throughout the completion of this thesis. Particularly, I am deeply thankful to my supervisor, Miguel Goulão, for his patience, guidance and systematic feedback throughout this process, as well as to my supervisor at Télécom Paris, Dominique Blouin, for his support and constant advice on the most varied topics. Their combined effort made this project possible and enriching.

I am equally grateful to NOVA FCT and Télécom Paris, whose partnership created this project and turned it into a life-changing opportunity. Being able to study and work between these two institutions was a unique and deeply enriching experience, both academically and personally.

I would also like to thank the colleagues I met thanks to this experience, who made it even more social and culturally fulfilling. Sharing this experience with them made each day more memorable and each challenge lighter.

My deepest gratitude goes to my parents, for their steady support, trust, and encouragement in every decision I make. Their belief in me gave me the confidence to go after opportunities such as this one. Without them, this would have never been possible.

I am also deeply grateful to my friends, to those who never forgot me, who were still there when I came back, and who went out of their way to meet me while I was away.

A very special thank you to my girlfriend, who pushed me to chase my goals and helped me along the way. Her encouragement, patience, and belief in me were a constant source of strength throughout this journey.

To everyone who believed in me and stood by my side, thank you.

% Last but not least, a sincere thank you to myself.

\end{ntacknowledgements}